---
title: "HW1"
format: pdf
---

# Homework 1

# Intro/Background

We are using an NIBRS-compliant [dataset](https://catalog.data.gov/dataset/crime
-data-from-2020-to-present) submitted to the FBI by the Los Angeles Police 
Department (LAPD). From the dataset website:
>This dataset reflects incidents of crime in the City of Los Angeles dating back
to 2020. This data is transcribed from original crime reports that are typed on 
paper and therefore there may be some inaccuracies within the data. Some 
location fields with missing data are noted as (0°, 0°). Address fields are only
provided to the nearest hundred block in order to maintain privacy. This data is
as accurate as the data in the database. Please note questions or concerns in 
the comments."

```{r, echo=FALSE}
library(ggplot2)
suppressPackageStartupMessages(library(dplyr))
#library(nnet)
library(reshape2)
#library(MatchIt)
suppressPackageStartupMessages(library(mgcv))
```

```{r, echo=FALSE}
library(readr)
crime_data <- suppressWarnings(
  read_csv("Crime_Data_from_2020_to_Present.csv",
                      locale = locale(tz = "America/Los_Angeles"),
                      show_col_types = FALSE)
)

#summary(crime_data)
```

## Examining the Data
```{r, echo=FALSE}
str(crime_data)
```

- This is an [official explanation](https://data.lacity.org/Public-Safety/Crime-
Data-from-2020-to-Present/2nrs-mtv8/about_data) of the columns from the LAPD


## Cleaning the Data

Initially the data has 986,500 rows corresponding to nearly 1 million individual
crimes. However, upon cleaning the data, namely by filtering out crimes where
the victim age is less than 0, if the victim sex is listed as "X" or "H", as
well as dropping rows where certain columns may be missing, we are left with a
dataset of 714,677 rows. Many of the aforementioned rows that have been dropped
correspond to crimes that do not have a specific victim such as fraud, etc. We
also rename the columns, replacing whitespace with underscores to avoid string
issues.

```{r, echo=FALSE}
# rename cols due ot issues with whitesapce
colnames(crime_data) <- gsub(" ", "_", colnames(crime_data))

# filter out rows where Vict Age is <= 0
crime_data <- crime_data %>% filter(`Vict_Age` > 0)

# Filter out rows where gender is "H" or "N/A"
crime_data <- crime_data %>%
    filter(!crime_data$`Vict_Sex` %in% c("X", "H") 
           & !is.na(`Vict_Sex`) & `Vict_Sex` != "")
    #filter(!crime_data$Status %in% c("CC"))

crime_data <- crime_data %>%
    filter(`Status` != "CC")

crime_data <- crime_data %>% 
    filter(!is.na(`Vict_Descent`))

invisible(sum(is.na(crime_data$Status)))
invisible(sum(is.na(crime_data$`Vict_Sex`)))
```

## Question 1 EDA

The first question we try to answer is whether the age of the victim 
(`Vict Age`) has a causal effect on the time a crime occurred (`Time Occ`).

We create bins to sort the age of victims into groups of 10 years each, so for
example 1-10, 11-20, etc. We also use bins to sort the time that the crime 
occured into 4 hour blocks.

```{r, echo=FALSE}

# convert TIME OCC to numeric, check for issues with parsing, and then remove 
# rows with issues
crime_data$`TIME_OCC` <- as.numeric(crime_data$`TIME_OCC`)
#problems(crime_data)
crime_data <- crime_data[!is.na(crime_data$`TIME_OCC`), ]

# wrap in invisible() so it doesnt print/output
# convert TIME OCC from HHMM to minutes since midnight
invisible(crime_data$time_in_minutes <- (crime_data$`TIME_OCC` %/% 100) * 60 
                              + (crime_data$`TIME_OCC` %% 100))

crime_data$time_in_minutes <- as.numeric(crime_data$time_in_minutes)

# str(crime_data$time_in_minutes)

# create bins for victim age
crime_data$vict_age_bins <- cut(crime_data$`Vict_Age`, 
                                breaks = seq(0, 100, by = 10), 
                                labels = paste(seq(0, 90, by = 10), 
                                               seq(10, 100, by = 10), 
                                               sep = "-"), 
                                include.lowest = TRUE)

# create bins for time (every 4 hours)
crime_data$time_bins <- cut(
  crime_data$time_in_minutes,
  breaks = seq(0, 1440, by = 240),
  labels = paste(
    sprintf("%02d:00", seq(0, 20, by = 4)),
    sprintf("%02d:00", seq(4, 24, by = 4)),
    sep = "-"
  ),
  include.lowest = TRUE
)
```

We then graph the counts of both the age and time bins, giving a general idea of
the count of crimes with respect to both age and time.

```{r, fig.height=6, echo=FALSE}
# crime count by age bins
age_bin_plot <- ggplot(crime_data, aes(x = vict_age_bins)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3) +
  labs(
    x = "Victim Age (10-year bins)",
    y = "Count of Incidents",
    title = "Crime Count by Victim Age Bins"
  )
  #theme_minimal()

age_bin_plot

# crime count by time bins
time_bin_plot <- ggplot(crime_data, aes(x = time_bins)) +
  geom_bar(fill = "darkorange") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3) +
  labs(
    x = "Time Occurred (4-hour bins)",
    y = "Count of Incidents",
    title = "Crime Count by Time Bins"
  ) +
  #theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

time_bin_plot
```

At a general glance, it seems that most victim ages are centered around 
individuals in their 20s to 40s. There does seem to be some crime for older
victims around 40-60. In terms of time bins, there doesn't seem to be as much
concentration, with a decent distribution of crime occurring throughout the day.
Crime incidents peak between 16:00-20:00 (4 PM to 8 PM), suggesting higher 
activity during the late afternoon and early evening. Also, there are fewer 
incidents between 04:00-08:00 (4 AM to 8 AM), which aligns with when people are
typically asleep. This distribution shows that crimes are more likely to occur 
during times when people are active and outside (afternoon and evening).


## Question 2 EDA

The second question we attempt to answer is whether the sex of the victim 
(`Vict Sex`) can affect the `Status` of the crime.

```{r, echo=FALSE}
ggplot(data = crime_data, aes(x = `Vict_Sex`, fill = `Vict_Sex`)) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.25) +
    scale_fill_manual(values = c("M" = "lightblue", "F" = "pink"))
```

```{r, fig.height=9.5, fig.width=10, echo=FALSE}
# Filter out rows where gender is "H" or "N/A"
crime_data <- crime_data %>%
      filter(!crime_data$`Vict_Sex` %in% c("X", "H") & !is.na(`Vict_Sex`) 
             & `Vict_Sex` != "")
      #filter(!crime_data$Status %in% c("CC"))

crime_data <- crime_data %>%
        filter(`Status` != "CC")

ggplot(data = crime_data, aes(x = `Vict_Sex`, fill = `Vict_Sex`)) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
    facet_wrap(~ `Status_Desc`, scales = "free_y") +
    scale_y_continuous() +
    labs(x = "Crime Status", y = "Count", title = "Victim Sex on Status") + 
    #theme_minimal() +
    scale_fill_manual(values = c("M" = "lightblue", "F" = "pink"))
```

Overall it seems that there are slightly more male victims of crime vs female
victims. That being said, it is kind of interesting that when it comes to 
juvenile arrests vs adult arrests, there are more male victims for the former,
whereas for the latter, adult arrests, there do seem to be more male arrests.
The juvenile and adult other categories basically act as a "catch-all", 
accounting for situations such as charges being dropped, citations, and other 
situations that did not lead to a conviction. It is also important to note there
is a significant increase in disparity when comparing Male and Female victims 
when it comes to investigation continuing.


# Analysis

## Question 1

To answer whether the age of the victim affects the time that the crime 
occurred, we decided to utilize a GAM without propensity score matching given 
that both victim age and time_in_minutes (from midnight), a custom column we 
made to be easier to work with, are continuous values. Overall, propensity score
matching would be much less impactful when not dealing with categorical values.
The GAM model includes the victim age as a smooth term to map non-linear 
relationships and adjusts for victim race, crime code, area, and victim sex as
additional covariate variables.

```{r, echo=FALSE}
# should work now after the weird stuff with whitespace
# fit GAM and set victim age as a smooth term for potential non-linear relation
age_time_model <- gam(time_in_minutes ~ s(`Vict_Age`) + `Vict_Descent` 
                      + `Crm_Cd` + `AREA` + `Vict_Sex`,
                data = crime_data, 
                family = gaussian())

```

## Question 2

Since we are dealing with a binary value, namely the victim being male or
female, we decided to utilize propensity score matching in order to compare
outcomes based on some covariate values. We used logistic regression via a GLM 
to model whether a victim as male or female based on the covariates. Then we 
we used nearest neighbor matching to match similar propensity scores in order to
create a matched dataset that reduces bias. The covariates in question we decided
to consider are the victim age bins, victim race, the crime (code), and the
general area the crime occurred in. We then used a multinomal logistic 
regression (given that there are more than 2 possible crime status outcomes) in 
order to model the (crime) Status variable based on victim sex. We did have to 
subset the data using a subset of 100,000 rows due to both memory and 
performance constraints. 

```{r, echo=FALSE}
# need to reload nnet due to conflicts with the mgcv GAM 
library(nnet)
library(MatchIt)

# set reference categ to "IC"
crime_data$Status <- as.factor(crime_data$Status)
crime_data$`Vict_Sex` <- as.factor(crime_data$`Vict_Sex`)
crime_data$`Status` <- relevel(crime_data$`Status`, ref = "IC")

# subset data due to memory issues
set.seed(123)
subset_crime_data <- crime_data[sample(nrow(crime_data), 100000), ]

# Propensity Score = estimate of being in a group (M or F)
# DO NOT include (Arrest) Status as covariate
propensity_model <- glm(`Vict_Sex` ~ `vict_age_bins` + `Vict_Descent` + `Crm_Cd` 
                        + `AREA`,
                        data = subset_crime_data, family = binomial)

propensity_scores <- predict(propensity_model, type = "response")

# nearest neighbor matching using propensity scores
matched_data <- matchit(`Vict_Sex` ~ `vict_age_bins` + `Vict_Descent` + `Crm_Cd` 
                        + `AREA`,
                        data = subset_crime_data, 
                        method = "nearest", 
                        ratio = 1, 
                        caliper = 0.05)

matched_data_df <- match.data(matched_data)

# need to use a multinomial logistic regression GLM since multiple categorical 
# variables for status, also some stuff with namespacing issues cause mgcv
sex_status_model <- nnet::multinom(`Status` ~ `Vict_Sex`, data = matched_data_df)
```


# Results

## Question 1

```{r, echo=FALSE}
summary(age_time_model)
```

Given that the p-value for the smooth term of victim age is extremely low, at 
p < 2e-16, there is a strong non-linear relationship between victim age and the
time that a crime occurred.

```{r, echo=FALSE}
# graphing this with plot vs ggplot is significantly less janky since it gives 
# predicted values with confidence intervals easily 
plot(age_time_model, 
     select = 1, 
     shade = TRUE, 
     shade.col = "lightblue", # confidence interval 
     col = "blue",
     se = TRUE, 
     main = "Effect of Victim Age on Crime Time", 
     ylab = "Estimated Effect on Crime Time (Minutes)", 
     xlab = "Victim Age",
     # make it a bit bigger for readability
     cex.main = 1.5,
     cex.lab = 1.2,
     cex.axis = 1)  

#ggplot(data = crime_data, aes(x = `Vict_Age`, ))
```

The solid blue line represents the fitted smooth function of victim age on the 
time a crime occurred. The shaded light blue area around the line represents a 
default 95% confidence interval, with a larger area corresponding to greater 
uncertainty at the respective age range, and vice-versa. In general, this plot 
shows how the estimated time of day for a crime tends to occur earlier or later 
depending on the victim’s age, with upward trends indicating crimes that occur
later in the day and downward trends indicating crimes that occur earlier in the
day for the respective age. (basically we look at the slope)

For example, when we look at younger victims, especially those closer to the age
of 20, it would appear that crimes involving them tend to occur much later in 
the day. However, when we get to the 20-30 age, it appears that the line 
approaches to 0, and from 30-40, the line remains relatively flat throughout the
age bracket, meaning that for those 20-30, crime occurs earlier in the day 
whereas for those 30-40, there isn't really any correlation to their age and the
time a crime occurs. For victims aged 40-90, it does seem like there is a 
gradual downward trend, indicating that crimes with older victims occur earlier
in the day. Interestingly enough, for those aged 90-100, the line trends up
again, indicating that crime related to them occur later in the day. However, 
the respective confidence interval area is also quite high, meaning that the
estimate is not as certain.

Overall, the age of a victim plays a significant role in determining when the 
crime in question occurs. 

## Question 2

```{r, echo=FALSE}
# coefficients give log odds
summary(sex_status_model)
```

In terms of the coefficients, they represent the log odds for each crime status
relative to the reference category, which in this case, we used a female victim
with the crime status being "IC" or investigation continuing. For example, when
the victim is male, then the log odds decrease by 0.1366 compared to if the 
victim is female for an adult arrest. Also, for juvenile arrests, then if the
victim is male, the log odds increases by 0.1371.

```{r, echo=FALSE}
# give odds ratios via exponentiation
exp(coef(sex_status_model))
```

Here we perform exponentiation on the coefficients to get the odds ratios in 
order to get values that are "easier" to interpret. For example, for an adult 
arrest, a value of 0.87 for a male victim means that an adult arrest for a male 
victim is 13% less likely compared to a female victim. For 0.745 for an AO, a 
male victim will be roughly 25.5% less likely to have an adult other status to a
crime vs a female victim. Also, for juvenile crimes, a male victim will be 14.7% 
more likely to lead to a juvenile arrest vs a female victim. In addition, a male
victim will be 37.6% more likely to have a juvenile other status.

```{r, echo=FALSE}
predicted_probs <- predict(sex_status_model, type = "probs")

predicted_probs_df <- as.data.frame(predicted_probs)
# need to add Victim Sex back into the DF
predicted_probs_df$`Vict Sex` <- matched_data_df$`Vict_Sex`

#head(predicted_probs_df)

# melt the data via reshape2 for easy ggplot graphing
predicted_probs_melt <- melt(predicted_probs_df, id.vars = "Vict Sex", 
                             variable.name = "Status", 
                             value.name = "Probability")

```

```{r, fig.height=5, echo=FALSE}
ggplot(predicted_probs_melt, aes(x = `Vict Sex`, y = Probability, 
                                 fill = `Vict Sex`)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = round(Probability, 4)), 
              vjust = -0.5) +
    facet_wrap(~ `Status`, scales = "free_y") +
    labs(title = "Predicted Probabilities of Arrest Status by Victim Sex",
       x = "Victim Sex", y = "Predicted Probability") +
    theme_minimal() +
    scale_fill_manual(values = c("M" = "lightblue", "F" = "pink")) +
    # use this for extra space above bars (so it doesnt cut off values)
    scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

Here we've decided to graph the respective predicted probability of both a male
and female victim of a crime being in each possible status. For example, lets 
look at the top middle graph, represented by "AA" or adult arrests. The female
bar has a value of 0.0911, and the male bar has a value of 0.0837. This means 
that 9.11% of male victim cases are expected to have an adult arrested status,
while 8.37% of female cases are expected to have an adult arrested status.

Overall it does seem that the sex of the victim does have some effect on the 
status of the crime. For example, if there is a female victim, there is a 
greater chance of the crime being an adult arrest, adult other, or juvenile
other. Conversely, if the victim is male, then there is a greater chance of the 
crime being an investigation continuing or juvenile arrest.


# Discussion

Regarding question 1 (does victim age affect crime time), its kind of 
interesting how in general, crime for older victims tends to occur earlier in 
the day, whereas for younger victims, crime occurs later in the day. A possible 
reason for this is that younger folks are either in school or working during the
day, whereas older folks could be out and about during the day. In addition, 
younger victims are more likely to be out during the evening, whereas older 
victims could be at home, sleeping early, etc. 

Regarding question 2 (does victim sex affect crime status), in general it seems
that for female victims, its much more likely for a crime to be "resolved" 
especially due to the disparity in male and female victims when considering how 
many more investigation continuing cases are comprised of male victims. It is 
important to note that at least for juvenile cases, male victims were more 
likely to lead to a juvenile arrest. Part of this could be related to the fact 
that there may be more public pressure to investigate and close cases that 
relate to violence against women.

In general, the dataset does seem somewhat limited. For example, there is no 
information about both the race and gender of the actual perpetrator of the 
crime. In addition, it would be great if there was a greater level of 
granularity when it came to describing crimes. For example, including whether a
crime as violent or non-violent, a sex-related crime, a felony or misdemeanor, 
etc would all be helpful in seeing more interesting trends. We think this is 
especially important as crime data can be interpreted in many ways and lead to 
serious real-world implications, such as additional funding, an increase in 
policing, and an overall varying approach towards how to even approach policing,
such as being stricter on crime vs a more harm-reduction, community based 
approach.
